---
title: "Portfolio 2"
author: "Anaelle Gackiere"
date: "02-13-2026"
output: html_document
---

### Project Description

The Substance Abuse and Mental Health Services Administration (SAMHSA)
pulbishes a yearly National Survey on Drug Use and Health (NSDUH). It is
an nationally representative, open-access, public-use dataset that is
designed to help researchers and clinicians gain a better understanding
of the state of mental health (MH) and substance used disorders (SUDs)
in America. Essentially, the NSDUH "is a nationwide survey that samples
the general population, 12 and older, on the use of tobacco, alcohol,
and drugs, mental health, and other health related issues in the United
States."

Although it is used primarily to monitor trends in MH and SUDs and
inform policies and resource allocation strategies. It also happens to
be a great data set to practice data cleaning, wrangling, visualization,
and analysis due to its large sample size, breadth of items, and strange
coding of variables.

A few notes before we dive in and explore this dataset: 1. Future
analyses will address sampling weights. 2. There are excluded personnel
in this dataset (prisoners, active-duty military members, etc...) 3.
Although I would love to have explored more variables and tested
specific hypotheses, this was not possible given the time frame. Future
work will explore specific theories. 4. You'll notice that I'm creating
many graphs, and looking at seemingly random variables and testing what
seems interesting. This is not good practice if the goal is inferential
testing. I am not worried at type 1 errors in this specific piece, given
it is mostly to explore a dataset I found to be interesting for
practice.

For more information on the variables used below, please see the
codebook attached in the p02 folder.

### Setting Up

First, I load all the libraries I think I might use (even if they go
unused, I prefer this over having to load the libraries on and on as I
go)

```{r load-data-libraries}
library(dplyr)      
library(ggplot2)    
library(survey)     
library(tidyverse)   
library(rstatix)  
library(lm.beta)
library(haven)
library(ggplot2)
library(ggridges)

# then we load the data
load("p02/NSDUH_2024.Rdata")

```

### Data Cleaning

If you have ever looked through this dataset, you'll notice a wide-range
of quirks that require some cleaning. Here, I'm interested in the
overall mental health score, and removing the responses that are skips
and such.

```{r cleaning}
# Clean mental health score
df <- NSDUH_2024 %>%
  mutate(
    MH_score = as.numeric(KSSLR6MONED),
    MH_score = ifelse(MH_score %in% c(94, 97, 98, 99), NA, MH_score)
  ) %>%
  # Remove rows with missing MH_score or AGE3
  filter(!is.na(MH_score), !is.na(AGE3))

```

Next, I want to look at young adults. The original entries are formatted
strangely (to me) and, although not best practice, I decide to lump them
into three big groups.

```{r age}
# Recode AGE3 into 3 broad adult age groups
df <- df %>%
  mutate(
    AGE_GROUP = case_when(
      AGE3 %in% 4:6  ~ "Young adult",   # 18–29
      AGE3 %in% 7:9  ~ "Middle age",    # 30–64
      AGE3 == 10 | AGE3 == 11 ~ "Older adult"
    ),
    AGE_GROUP = factor(AGE_GROUP, levels = c("Young adult", "Middle age", "Older adult"))
  )

# Check counts
table(df$AGE_GROUP)
```

### Question 1: Is age group related to mental health scores?

```{r sig-differences}
# let's plot it!
ggplot(df, aes(x = MH_score, fill = AGE_GROUP)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Mental Health Score",
    y = "Density",
    fill = "Age Group",
    title = "Mental Health Scores by Age Group"
  ) +
  theme_minimal()

# I love inferential testing!! 

# Kruskal-Wallis test --> using this since our data is very skewed
kruskal.test(MH_score ~ AGE_GROUP, data = df)

# Effect size
kruskal_effsize(df, MH_score ~ AGE_GROUP)

# Pairwise Wilcoxon comparisons
pairwise.wilcox.test(df$MH_score, df$AGE_GROUP, p.adjust.method = "bonferroni")
```

Conclusion: There is a relationship between age group and mental health
scores. Young adults have the higher mental health scores than middle
aged and older adults. However, the distribution is highly skewed for
all three groups (which is why I used a Kruskal-Wallist test).

### Data Cleaning and Discovery Continued

The young adult data stood out to me in the graph above. I wanted to
explore how these young adults are doing in more depth.

As I was going through the codebook, a few variables stood out to me.
Here, I'm cleaning them just like I did for the mental health scores,
and I'm renaming them for clarity. I'm also creating a new df with just
the young adult age group. You'll notice I did not use every variable,
but I wanted to clean them nonetheless in case I might use them.

```{r young-adults}
# lots of cleaning to do
young_adults <- df %>%
  filter(AGE_GROUP == "Young adult")

young_adults <- young_adults %>%
  mutate(
    POVERTY3   = ifelse(POVERTY3 %in% c(94, 97, 98, 99), NA, POVERTY3),
    HRTCONDEV2 = ifelse(HRTCONDEV2 %in% c(94, 97, 98, 99), NA, HRTCONDEV2),
    BMI2       = ifelse(BMI2 %in% c(94, 97, 98, 99), NA, BMI2),
    CASURCVR   = ifelse(CASURCVR %in% c(94, 97, 98, 99), NA, CASURCVR),
    CAMHPROB   = ifelse(CAMHPROB %in% c(94, 97, 98, 99), NA, CAMHPROB),
    HEALTH2    = ifelse(HEALTH2 %in% c(94, 97, 98, 99), NA, HEALTH2),
    IRSEX      = ifelse(IRSEX %in% c(94, 97, 98, 99), NA, IRSEX),
    HLCNOTYR   = ifelse(HLCNOTYR %in% c(94, 97, 98, 99), NA, HLCNOTYR),
    GOVTPROG   = ifelse(GOVTPROG %in% c(94, 97, 98, 99), NA, GOVTPROG)
  ) %>%
  rename(
    poverty_level = POVERTY3,
    heart_condition = HRTCONDEV2,
    bmi = BMI2,
    drug_problem = CASURCVR,
    mental_health_problem = CAMHPROB,
    overall_health = HEALTH2,
    gender = IRSEX,
    no_health_insurance = HLCNOTYR,
    gov_program = GOVTPROG
  ) %>%
   # Convert categorical predictors to factors
  mutate(
    heart_condition = factor(heart_condition),
    drug_problem = factor(drug_problem),
    mental_health_problem = factor(mental_health_problem),
    gender = factor(gender),
    no_health_insurance = factor(no_health_insurance),
    gov_program = factor(gov_program)
  ) %>%
  # Remove any rows with missing values in DV or predictors
  drop_na(MH_score, poverty_level, heart_condition, bmi, drug_problem,
          mental_health_problem, overall_health, gender, no_health_insurance, gov_program)

# now we save this
young_adults <- young_adults %>%
  select(MH_score, poverty_level, heart_condition, bmi, drug_problem,
         mental_health_problem, overall_health, gender, no_health_insurance, gov_program)

```

Now that we're focusing on the young adults, I want to see what the
distribution, in the form of a histogram, looks like for mental health
scores in that group.

```{r health}
# let's take a peak
head(young_adults)
 
summary(young_adults)

# this loks pretty normally distributed among young people
ggplot(young_adults, aes(MH_score)) +
  geom_histogram(bins = 30, color = "plum4", fill = "plum")

```

### Question 2: To what degree is overall heatlh related to mental health in young adults?

I want to look at overall health and mental health. I'm reverse scoring
overall health so that 4 means great health and 1 means poor health.
Then, I visualize that relationship between mental health and overall
health with a jitter plot.

```{r health2}
# mutate so it makes more sense
young_adults <- young_adults %>%
  mutate(overall_health_rev = recode(overall_health,
                                     `1` = 4,
                                     `2` = 3,
                                     `3` = 2,
                                     `4` = 1
  ))

# relationship btwn overall and MH 
ggplot(young_adults, aes(factor(overall_health_rev), MH_score)) +
  geom_jitter(aes(colour = overall_health_rev), width = .15, alpha = .3) +
  labs(title = "Overall Health vs Mental Health",
       subtitle = "young adults: age 18-29",
       )

```

Our visualization demonstrates that there is in fact a relationship
between overall health and mental health in young adults (which should
not come as a surprise)

### Data Cleaning Continued

Before we visualize and analyze more variables, I need to understand
what class each variable is in, as that will help me when creating
certain plots. I'm tidying up the data now and making it longer (which
decreases the amount of columns).

```{r}
# which type of var
sapply(young_adults, class)

newDF <- young_adults %>%
  filter(bmi < 45) %>%
  select(MH_score, no_health_insurance, gov_program, gender, overall_health_rev) %>%
  mutate(overall_health_rev = factor(overall_health_rev, levels = 1:4,
                                     labels = c("Poor", "Fair", "Good", "Excellent"))) %>%
  pivot_longer(-MH_score, names_to = "predictor", values_to = "group")
```

### Data Visualization

Now, you'll see density ridges for various variables. This helps me
determine if anything stands out and might be interesting to analyze
more closely.

```{r overall-health}
ggplot(newDF, aes(x = MH_score, y = group)) +
  geom_density_ridges2() +
  facet_wrap(~ predictor, scales = "free_y") +
  labs(y = "Group", x = "Mental Health Score")

```

### Question 3: What's the relationship between Mental Health Scores and BMI in young adults?

The BMI distribution really stood out to me. First, I'm getting rid of
BMI over 45, given that anything over that point may be a typo (if this
were a real data analysis project, I would investigate this more
carefully before ommitting these cases).

```{r}
# the bmi one also really stood out to me

young_adults_bmi <- young_adults %>% filter(bmi < 45)

ggplot(young_adults_bmi, aes(x = bmi, y = MH_score)) +
  geom_point(alpha = .25) +
  geom_smooth(method = "loess") +
  labs(x = "BMI", y = "Mental Health Score")

# 
young_adults_bmi %>%
  mutate(bmi_cat = cut(bmi, breaks = 4)) %>%
  ggplot(aes(x = bmi_cat, y = MH_score)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = .2) +
  labs(x = "BMI category", y = "Mental Health Score")


# is there really a u shaped curve? - yes
lm(MH_score ~ bmi + I(bmi^2), data = young_adults_bmi)

ggplot(young_adults_bmi, aes(x = bmi, y = MH_score)) +
  geom_point(alpha = .2) +
  geom_smooth(method = "gam", formula = y ~ s(x), size = 1.2) +
  theme_minimal() +
  labs(
    title = "Curvilinear Association Between BMI and Mental Health",
    x = "Body Mass Index (BMI)",
    y = "Mental Health Score"
  )
```

At first, I found a curvilinear association. I wanted to verify whether
that was really the case. I conducted a regression and plotted it again.
It is in fact curvilinear! My suspicions are that individuals with
higher BMIs likely face discrimination based on their weight, which can
impact their mental health. Since obesity and weight gain is
multifactorial, it is likely that factors like poverty, low social
support, food deserts, and such contribute to this association as well.
For individuals with lower BMI, this association may be due to body
dysmorphia, eating disorders with severe weightloss involved, or other
health issues that lead to weight loss.

### Question 4: Is there a difference in BMI vs MH depending on gender?

```{r cool stuff}
# Question 4
ggplot(young_adults_bmi,
       aes(x = bmi, y = MH_score, color = factor(overall_health_rev))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = FALSE) +
  labs(
    x = "BMI",
    y = "Mental Health Score",
    color = "Overall Health",
    title = "BMI vs MH, colored by Overall Health"
  ) +
  theme_minimal()


# Plot BMI vs Mental Health by gender
ggplot(young_adults_bmi, aes(x = bmi, y = MH_score, color = gender)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = FALSE) +
  theme_minimal() +
  labs(
    title = "BMI vs Mental Health by Gender",
    x = "Body Mass Index (BMI)",
    y = "Mental Health Score",
    color = "Gender"
  )


```

Here's what I've found so far: Among young adults, mental health is not
linearly related to BMI; instead, both very low and very high BMI are
associated with worse mental health, and this relationship varies by
overall health and gender.

### Modeling Gender

```{r model}

# Gender seemed interesting so lets model that. 

model <- lm(
  MH_score ~ bmi + I(bmi^2) + gender + overall_health_rev +
             poverty_level + no_health_insurance,
  data = young_adults_bmi
)

summary(model)
lm.beta(model)

```

BMI and overall health have a meaningful effects (curvilenar). Lack of
health insurance was not significant in this model, but this may be due
to the fact that the young adults in our sample have access to health
insurance, and may not be representative of the population.

But let's look at gender moderation:

```{r gender}
lm(
  MH_score ~ bmi + I(bmi^2)*gender + overall_health_rev +
             poverty_level,
  data = young_adults_bmi
)

```

## Conclusion

overall health MAJOR predictor.Is the BMI–mental health relationship
different for men and women? Yes. The main effect seems weak but there
is definitely moderation.

so: Although gender was not a strong predictor of average mental health,
it significantly moderated the curvilinear association between BMI and
mental health, indicating that body-related mental health risks operate
differently across genders.

**References**: Center for Behavioral Health Statistics and Quality.
(2025). 2024 National Survey on Drug Use and Health Public Use File
Codebook, Substance Abuse and Mental Health Services Administration,
Rockville, MD
