---
title: "Portfolio 4"
output:
  html_document: 
    theme: yeti
    highlight: zenburn
  word_document: default
date: "2026-03-20"
---

# **Spotify Data Analysis**

In this portfolio, I decided to analyze my spotify data to learn more about my musical preferences.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(jsonlite)
library(lubridate)
library(scales)
```

## Loading in my spotify data

```{r load-data}
json_files <- list.files(path = "p04/Spotify Account Data", 
                         pattern = "StreamingHistory_music.*\\.json",
                         full.names = TRUE)

raw <- json_files %>%
  map(fromJSON) %>%
  bind_rows()

# Preview the raw data
glimpse(raw)
```

## Cleaning the Data

```{r clean-data}
df <- raw %>%
  rename(
    timestamp = endTime,
    ms_played = msPlayed,
    track     = trackName,
    artist    = artistName
  ) %>%
  mutate(
    timestamp   = ymd_hm(timestamp),  # note: ymd_hm not ymd_hms, no seconds in basic format
    date        = as_date(timestamp),
    hour        = hour(timestamp),
    day_of_week = wday(timestamp, label = TRUE, abbr = FALSE),
    month       = floor_date(date, "month"),
    minutes_played = ms_played / 60000
  ) %>%
  filter(
    !is.na(track),
    ms_played > 30000
  )
```

## Overview

```{r overview}
cat("Total plays:", nrow(df), "\n")
cat("Unique tracks:", n_distinct(df$track), "\n")
cat("Unique artists:", n_distinct(df$artist), "\n")
cat("Date range:", as.character(min(df$date)), "to", as.character(max(df$date)), "\n")
cat("Total hours listened:", round(sum(df$minutes_played) / 60, 1), "\n")
```

## Top Artists

```{r top-artists, fig.width=8, fig.height=5}
df %>%
  count(artist, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(artist = fct_reorder(artist, n)) %>%
  ggplot(aes(x = n, y = artist)) +
  geom_col(fill = "plum4") +
  labs(
    title = "My Top 15 Most Played Artists",
    x = "Number of Plays",
    y = NULL
  ) +
  theme_minimal(base_size = 13)
```

## Top Tracks

```{r top-tracks, fig.width=8, fig.height=5}
df %>%
  count(track, artist, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(label = paste0(track, " — ", artist),
         label = fct_reorder(label, n)) %>%
  ggplot(aes(x = n, y = label)) +
  geom_col(fill = "plum4") +
  labs(
    title = "My Top 15 Most Played Songs",
    x = "Number of Plays",
    y = NULL
  ) +
  theme_minimal(base_size = 13)
```

## Listening by Time of Day

```{r time-of-day, fig.width=8, fig.height=4}
df %>%
  count(hour) %>%
  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "plum4") +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "When Do I Listen to Music?",
    x = "Hour of Day (0 = midnight)",
    y = "Number of Plays"
  ) +
  theme_minimal(base_size = 13)
```

## Listening by Day of Week

```{r day-of-week, fig.width=8, fig.height=4}
df %>%
  count(day_of_week) %>%
  ggplot(aes(x = day_of_week, y = n)) +
  geom_col(fill = "plum4") +
  labs(
    title = "Which Days Do I Listen Most?",
    x = NULL,
    y = "Number of Plays"
  ) +
  theme_minimal(base_size = 13)
```

## Listening Over Time

```{r over-time, fig.width=9, fig.height=4}
df %>%
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_line(color = "plum", linewidth = 1.2) +
  geom_point(color = "plum4", size = 2) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "My Listening Activity Over the Past Year",
    x = NULL,
    y = "Number of Plays"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Minutes Listened per Month

```{r minutes-per-month, fig.width=9, fig.height=4}
df %>%
  group_by(month) %>%
  summarise(hours = sum(minutes_played) / 60) %>%
  ggplot(aes(x = month, y = hours)) +
  geom_col(fill = "plum4") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "Hours Listened per Month",
    x = NULL,
    y = "Hours"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Artist Diversity Over Time

How many unique artists did I listen to each month?

```{r artist-diversity, fig.width=9, fig.height=4}
df %>%
  group_by(month) %>%
  summarise(unique_artists = n_distinct(artist)) %>%
  ggplot(aes(x = month, y = unique_artists)) +
  geom_line(color = "plum", linewidth = 1.2) +
  geom_point(color = "plum4", size = 2) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "How Many Different Artists Did I Listen to Each Month?",
    x = NULL,
    y = "Unique Artists"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Notes

Data covers the past year of streaming history from Spotify's data export. Plays under 30 seconds were excluded as likely skips and podcasts were excluded (no track metadata).

### Saving data for the shiny app

```{r}
saveRDS(df, "p04/spotify_data.rds")
```

**To view the Shiny App, go to p04 folder, and you'll find it in the Spotify folder!**
