---
title: "Portfolio 9"
author: "Anaelle Gackiere"
date: "`r Sys.Date()`"
---


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/anaellegackiere/Desktop/R_Lab/R-Portfolio/p09")

library(hBayesDM)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(tidyverse)
library(R.matlab)
library(pastecs)
library(rstatix)
library(emmeans)
library(afex)
library(effectsize)
library(ggpubr)
library(readxl)
library(knitr)
library(kableExtra)
library(interactions)
library(mediation)
library(loo)
library(rstan)
```

---

# Overview

This report presents the analyses for the **Time Task** of the Social Engagement Reinforcement Learning paradigm. Participants provided ratings of how different amounts of seconds spent with someone else would motivate them to participate in a Jenga activity (+15 secs, 0 secs, -15 secs). They also rated how much time, if given the choice, they would like to spend playing with another student. Afterwards, they completed a reinforcement learning task where outcomes varied in how much time they spent with other people (same as explicit ratings: positive = + 15 secs, neutral = 0 seconds, negative = - 15 seconds).

The measure of interest is the **computationally inferred Reward Value (cRV / k parameter)**, which captures the reward values each time outcome through their behavior.

---

# Setup

```{r time-setup}
load("preprocessed_Time_behavior.RData")
load("analysis_ready_Time.RData")
load("loo_objects_Time.RData")

# Load k parameters
parT <- readMat("Qparams_ABK_FYP_Time.mat")
N_T = length(parT$alpha.QabkF[1,])

alphaT = vector(length = N_T)
betaT  = vector(length = N_T)
kHiT   = vector(length = N_T)
kLowT  = vector(length = N_T)
kMedT  = vector(length = N_T)

for (i in 1:N_T) {
  alphaT[i] <- median(parT$alpha.QabkF[,i])
  betaT[i]  <- median(parT$beta.QabkF[,i])
  kHiT[i]   <- median(parT$kPos.QabkF[,i])
  kLowT[i]  <- median(parT$kNeg.QabkF[,i])
  kMedT[i]  <- median(parT$kNeu.QabkF[,i])
}
```

---

# Explicit Ratings

## Descriptive Statistics

```{r time-explicit-desc}
desc_table_T <- data.frame(
  Condition = c("Positive (more time)", "Neutral", "Negative (less time)"),
  N         = c(sum(!is.na(analysis_df_time$PosTime_1)),
                sum(!is.na(analysis_df_time$NeutralTime_1)),
                sum(!is.na(analysis_df_time$NegTime_1))),
  Mean      = round(c(mean(analysis_df_time$PosTime_1, na.rm=TRUE),
                      mean(analysis_df_time$NeutralTime_1, na.rm=TRUE),
                      mean(analysis_df_time$NegTime_1, na.rm=TRUE)), 3),
  SD        = round(c(sd(analysis_df_time$PosTime_1, na.rm=TRUE),
                      sd(analysis_df_time$NeutralTime_1, na.rm=TRUE),
                      sd(analysis_df_time$NegTime_1, na.rm=TRUE)), 3)
)

kable(desc_table_T, caption = "Descriptive Statistics - Explicit Ratings (Time Task)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## T-tests

```{r time-explicit-ttests}
te1 <- t.test(analysis_df_time$PosTime_1, analysis_df_time$NeutralTime_1, paired=TRUE)
te2 <- t.test(analysis_df_time$PosTime_1, analysis_df_time$NegTime_1, paired=TRUE)
te3 <- t.test(analysis_df_time$NeutralTime_1, analysis_df_time$NegTime_1, paired=TRUE)

ttest_table_T <- data.frame(
  Comparison = c("Positive vs Neutral", "Positive vs Negative", "Neutral vs Negative"),
  t          = round(c(te1$statistic, te2$statistic, te3$statistic), 3),
  df         = c(te1$parameter, te2$parameter, te3$parameter),
  p          = round(c(te1$p.value, te2$p.value, te3$p.value), 4),
  Mean_diff  = round(c(te1$estimate, te2$estimate, te3$estimate), 3)
)

kable(ttest_table_T, caption = "Paired T-tests - Explicit Ratings (Time Task)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Plot

```{r time-explicit-plot, fig.width=7, fig.height=5}
ePosT <- data.frame(value = analysis_df_time$PosTime_1,     param = "Positive")
eNeuT <- data.frame(value = analysis_df_time$NeutralTime_1, param = "Neutral")
eNegT <- data.frame(value = analysis_df_time$NegTime_1,     param = "Negative")
eAllT <- rbind(ePosT, eNeuT, eNegT)
eAllT$param <- factor(eAllT$param, levels = c("Positive", "Neutral", "Negative"))

ggplot(eAllT, aes(x = param, y = value, fill = param)) +
  geom_violin(position = position_dodge(0.9)) +
  geom_boxplot(width = 0.2, position = position_dodge(0.9)) +
  labs(x = "Condition", y = "Explicit Rating", title = "Explicit Ratings by Condition (Time Task)") +
  theme_classic() +
  theme(legend.position = "none")
```

---

# Implicit Motivation (cRV / k parameters)

## Descriptive Statistics

```{r time-imv-desc}
imv_desc_T <- data.frame(
  Condition = c("Positive/More Time (kPos)", "Negative/Less Time (kNeg)", "Neutral (kNeu)"),
  N         = c(length(kHiT), length(kLowT), length(kMedT)),
  Mean      = round(c(mean(kHiT), mean(kLowT), mean(kMedT)), 3),
  SD        = round(c(sd(kHiT), sd(kLowT), sd(kMedT)), 3),
  Median    = round(c(median(kHiT), median(kLowT), median(kMedT)), 3)
)

kable(imv_desc_T, caption = "Descriptive Statistics - IMV Parameters (Time Task)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## T-tests

```{r time-imv-ttests}
ti1_T <- t.test(analysis_df_time$PosTime_imv, analysis_df_time$NegTime_imv, paired=TRUE)
ti2_T <- t.test(analysis_df_time$PosTime_imv, analysis_df_time$NeuTime_imv, paired=TRUE)
ti3_T <- t.test(analysis_df_time$NeuTime_imv, analysis_df_time$NegTime_imv, paired=TRUE)

imv_ttest_T <- data.frame(
  Comparison = c("Positive vs Negative", "Positive vs Neutral", "Neutral vs Negative"),
  t          = round(c(ti1_T$statistic, ti2_T$statistic, ti3_T$statistic), 3),
  df         = c(ti1_T$parameter, ti2_T$parameter, ti3_T$parameter),
  p          = round(c(ti1_T$p.value, ti2_T$p.value, ti3_T$p.value), 4),
  Mean_diff  = round(c(ti1_T$estimate, ti2_T$estimate, ti3_T$estimate), 3)
)

kable(imv_ttest_T, caption = "Paired T-tests - IMV Parameters (Time Task)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Effect Size (Positive vs Negative)

```{r time-imv-effectsize}
effectsize(t.test(analysis_df_time$PosTime_imv, analysis_df_time$NegTime_imv, var.equal=TRUE, paired=TRUE))
```

## Plot

```{r time-imv-plot, fig.width=7, fig.height=5}
kHiTDat  <- data.frame(value = kHiT, param = "Positive")
kLowTDat <- data.frame(value = kLowT, param = "Negative")
kMidTDat <- data.frame(value = kMedT, param = "Neutral")
HiLowT   <- rbind(kHiTDat, kMidTDat, kLowTDat)
HiLowT$param <- factor(HiLowT$param, levels = c("Positive", "Neutral", "Negative"))

ggplot(HiLowT, aes(x = param, y = value, fill = param)) +
  geom_violin(position = position_dodge(0.9)) +
  geom_boxplot(width = 0.2, position = position_dodge(0.9)) +
  labs(x = "Condition", y = "cRV (k)", title = "Implicit Motivation by Condition (Time Task)") +
  theme_classic() +
  theme(legend.position = "none")
```

---

# Implicit-Explicit Correlations

## Full Difference Scores

```{r time-imp-exp-full}
c_full <- cor.test(analysis_df_time$PosvsNegTime_diff_imv, analysis_df_time$PosvsNegTime_diff_explicit)

kable(data.frame(
  Comparison = "IMV diff (Pos-Neg) ~ Explicit diff (Pos-Neg)",
  r          = round(c_full$estimate, 3),
  p          = round(c_full$p.value, 4),
  CI_lower   = round(c_full$conf.int[1], 3),
  CI_upper   = round(c_full$conf.int[2], 3)
), caption = "Full Difference Score Correlation") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Sub-Difference Scores

```{r time-imp-exp-sub}
cs1 <- cor.test(analysis_df_time$PosNeu_diff_imv, analysis_df_time$PosTime_1 - analysis_df_time$NeutralTime_1)
cs2 <- cor.test(analysis_df_time$NeuNeg_diff_imv, analysis_df_time$NeutralTime_1 - analysis_df_time$NegTime_1)

sub_cor_table <- data.frame(
  Comparison = c("IMV Pos-Neu ~ Explicit Pos-Neu", "IMV Neu-Neg ~ Explicit Neu-Neg"),
  r          = round(c(cs1$estimate, cs2$estimate), 3),
  p          = round(c(cs1$p.value, cs2$p.value), 4),
  CI_lower   = round(c(cs1$conf.int[1], cs2$conf.int[1]), 3),
  CI_upper   = round(c(cs1$conf.int[2], cs2$conf.int[2]), 3)
)

kable(sub_cor_table, caption = "Sub-Difference Score Correlations: Implicit ~ Explicit") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(sub_cor_table$p < .05), bold = TRUE, color = "white", background = "#2c7bb6")
```

---

# Social Preference Rating Analyses

## Social vs. Solo Ratings 

```{r}
# Social vs Solo ratings plot
social_solo_dat <- data.frame(
  value = c(as.numeric(analysis_df_time$Social_Rating), 
            as.numeric(analysis_df_time$Solo_Rating)),
  param = c(rep("Social", nrow(analysis_df_time)), 
            rep("Solo", nrow(analysis_df_time)))
)

social_solo_dat$param <- factor(social_solo_dat$param, levels = c("Social", "Solo"))

ggplot(social_solo_dat, aes(x = param, y = value, fill = param)) +
  geom_boxplot(width = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  labs(x = "Condition", y = "Rating",
       title = "Social vs Solo Task Preference Ratings") +
  theme_classic() +
  theme(legend.position = "none")

t.test(as.numeric(analysis_df_time$Social_Rating),
       as.numeric(analysis_df_time$Solo_Rating),
       paired = TRUE)
```

## Correlations with Rating Diff (Social - Solo)

```{r time-rating-cors}
rct1 <- cor.test(analysis_df_time$PosvsNegTime_diff_imv, analysis_df_time$Rating_diff)
rct2 <- cor.test(analysis_df_time$PosvsNegTime_diff_explicit, analysis_df_time$Rating_diff)
rct3 <- cor.test(analysis_df_time$PosNeu_diff_imv, analysis_df_time$Rating_diff)
rct4 <- cor.test(analysis_df_time$NeuNeg_diff_imv, analysis_df_time$Rating_diff)

rating_cor_T <- data.frame(
  Predictor = c("IMV diff (Pos-Neg)", "Explicit diff (Pos-Neg)",
                "IMV diff (Pos-Neu)", "IMV diff (Neu-Neg)"),
  r         = round(c(rct1$estimate, rct2$estimate, rct3$estimate, rct4$estimate), 3),
  p         = round(c(rct1$p.value, rct2$p.value, rct3$p.value, rct4$p.value), 4),
  CI_lower  = round(c(rct1$conf.int[1], rct2$conf.int[1], rct3$conf.int[1], rct4$conf.int[1]), 3),
  CI_upper  = round(c(rct1$conf.int[2], rct2$conf.int[2], rct3$conf.int[2], rct4$conf.int[2]), 3)
)

kable(rating_cor_T, caption = "Correlations with Social-Solo Rating Difference (Time Task)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(rating_cor_T$p < .05), bold = TRUE, color = "white", background = "#2c7bb6")
```


---

# Trait Correlations

```{r time-trait-cors}
trait_vars_T <- c("Extraversion", "Neuroticism", "Openness", "LSAS",
                  "Prosocial_Interactions", "Neg_Social_Potency", "Sociability",
                  "Admiration", "Passivity", "Sexual_Relations",
                  "Time_Actual", "Time_Choice_1")

dv_vars_T <- c("PosvsNegTime_diff_imv", "PosvsNegTime_diff_explicit")

cor_table_fn <- function(df, dv_vars, trait_vars) {
  results <- list()
  for (dv in dv_vars) {
    dv_results <- lapply(trait_vars, function(trait) {
      res <- cor.test(df[[dv]], df[[trait]], use = "complete.obs")
      data.frame(
        DV       = dv,
        Trait    = trait,
        r        = round(as.numeric(res$estimate), 3),
        p        = round(res$p.value, 3),
        CI_lower = round(res$conf.int[1], 3),
        CI_upper = round(res$conf.int[2], 3)
      )
    })
    results[[dv]] <- do.call(rbind, dv_results)
  }
  do.call(rbind, results)
}

cor_results_T <- cor_table_fn(analysis_df_time, dv_vars_T, trait_vars_T)

kable(subset(cor_results_T, DV == "PosvsNegTime_diff_imv")[,-1],
      caption = "IMV Diff ~ Traits (Time Task)", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(subset(cor_results_T, DV == "PosvsNegTime_diff_imv")$p < .05),
           bold = TRUE, color = "white", background = "#2c7bb6")

kable(subset(cor_results_T, DV == "PosvsNegTime_diff_explicit")[,-1],
      caption = "Explicit Diff ~ Traits (Time Task)", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(subset(cor_results_T, DV == "PosvsNegTime_diff_explicit")$p < .05),
           bold = TRUE, color = "white", background = "#2c7bb6")
```

---


# Moderation Analyses

## Does Extraversion Moderate the Implicit-Explicit Link?

```{r time-mod-extra}
mod_extra_impexp <- lm(PosvsNegTime_diff_explicit ~ imv_diff_c * extraversion_c, 
                       data = analysis_df_time)
summary(mod_extra_impexp)
```
*IMV diff is a strong predictor (p < .001). People with higher implicit valuation 
differentiation also show higher explicit valuation differentiation. Extraversion is 
a significant main effect (p = .018). More extraverted people show larger explicit 
valuation differences between positive and negative time conditions. The interaction 
is significant (p = .033). Extraversion moderates the implicit-explicit relationship. 
The negative coefficient suggests the link between implicit and explicit valuation is 
stronger for less extraverted (more introverted) people.* 

```{r}
# Explicit differentiation by extraversion
ggplot(analysis_df_time, aes(x = Extraversion, y = PosvsNegTime_diff_explicit)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "#2c7bb6") +
  labs(x = "Extraversion", y = "Explicit Diff (Pos - Neg)",
       title = "Extraversion Predicts Explicit Social Time Differentiation") +
  theme_classic()

# IMV differentiation by extraversion
ggplot(analysis_df_time, aes(x = Extraversion, y = PosvsNegTime_diff_imv)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "#2c7bb6") +
  labs(x = "Extraversion", y = "IMV Diff (Pos - Neg)",
       title = "Extraversion Predicts Implicit Social Time Differentiation") +
  theme_classic()

```


---

# State Social Motivation: Time Choice → IMV

**Time_Choice_1** is a state/motivational measure (how much do I want to be with people right now), whereas **Extraversion** is a trait measure (how much do I generally want to be with people). The RL measure appears sensitive to state-level motivation.

## Time Choice Predicts IMV Diff

```{r time-choice-imv}
mod_timechoice <- lm(PosvsNegTime_diff_imv ~ time_choice_c, data = analysis_df_time)
summary(mod_timechoice)
```

*Prior explicit social motivation (Time Choice) strongly predicts implicit learning differentiation (R² = .24, p < .001) — people who want to spend more time with others show greater implicit valuation of positive time outcomes.*

## Does Extraversion Moderate the Time Choice → IMV Link?

```{r time-choice-imv-mod}
mod_timechoice_extra <- lm(PosvsNegTime_diff_imv ~ time_choice_c * extraversion_c, 
                           data = analysis_df_time)
summary(mod_timechoice_extra)
```

*The Time Choice → IMV relationship is equally strong regardless of extraversion level (interaction p = .572), suggesting the RL measure captures state-level social motivation independently of stable personality traits.*


```{r}
# Histogram of Time Choice
ggplot(analysis_df_time, aes(x = Time_Choice_1)) +
  geom_histogram(binwidth = 1, fill = "#2c7bb6", color = "white", alpha = 0.8) +
  labs(x = "Time Choice (minutes)", y = "Count",
       title = "Distribution of Chosen Time with Others") +
  theme_classic()

# Time Choice ~ IMV diff
ggplot(analysis_df_time, aes(x = Time_Choice_1, y = PosvsNegTime_diff_imv)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "#2c7bb6") +
  labs(x = "Time Choice (minutes)", y = "IMV Diff (Pos - Neg)",
       title = "Time Choice vs. cRVs") +
  theme_classic()

cor.test(analysis_df_time$Time_Choice_1, analysis_df_time$PosvsNegTime_diff_imv)

# Time Choice ~ Explicit diff
ggplot(analysis_df_time, aes(x = Time_Choice_1, y = PosvsNegTime_diff_explicit)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "#2c7bb6") +
  labs(x = "Time Choice (minutes)", y = "Explicit Diff (Pos - Neg)",
       title = "Time Choice vs. Explicit Ratings") +
  theme_classic()

cor.test(analysis_df_time$Time_Choice_1, analysis_df_time$PosvsNegTime_diff_explicit)
```

Time Choice) predicts both implicit and explicit differentiation, but the relationship is stronger for the implicit measure.




```{r}
# Create comprehensive correlation table for all IMV and explicit measures
all_vars <- c("PosTime_imv", "NegTime_imv", "NeuTime_imv", "PosvsNegTime_diff_imv",
              "PosNeu_diff_imv", "NeuNeg_diff_imv",
              "PosTime_1", "NegTime_1", "NeutralTime_1", "PosvsNegTime_diff_explicit",
              "Rating_diff", "Time_Choice_1", "Time_Actual",
              "Extraversion", "Neuroticism", "LSAS", "Prosocial_Interactions",
              "Sociability", "Social_Rating", "Solo_Rating")

# Build correlation matrix
cor_matrix <- cor(analysis_df_time[, all_vars], use = "pairwise.complete.obs")
p_matrix <- matrix(NA, nrow = length(all_vars), ncol = length(all_vars))
for (i in 1:length(all_vars)) {
  for (j in 1:length(all_vars)) {
    if (i != j) {
      p_matrix[i,j] <- cor.test(analysis_df_time[[all_vars[i]]], 
                                  analysis_df_time[[all_vars[j]]])$p.value
    }
  }
}

rownames(p_matrix) <- colnames(p_matrix) <- all_vars

# Display as kable - round and mark significant ones
cor_display <- round(cor_matrix, 2)
cor_display[p_matrix > .05] <- NA  # blank out non-significant

kable(cor_display, 
      caption = "Correlation Matrix (significant correlations only, p < .05)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                font_size = 11)
```



# Model Comparison
```{r model-comparison}
# MODEL COMPARISON

# Load saved LOO objects
load("loo_objects_Time.RData")

# Load models for WAIC
load("fitQ_ABK_FYP_Time")
load("fitQ_explicit_Time.RData")

# Extract log likelihoods
loglik_Q_full <- extract_log_lik(fitQabkF, merge_chains = FALSE)
loglik_explicit <- extract_log_lik(fitQ_explicit, merge_chains = FALSE)

# Subset k model to same participants as explicit model
na_idx <- is.na(as.numeric(analysis_df_time$PosTime_1[match(
  read.csv("ParticipantIDs_Time.csv")$participantIDs, analysis_df_time$PID)]))

# WAIC
waic_k        <- waic(loglik_Q_full[,,!na_idx])
waic_explicit <- waic(loglik_explicit)

# LOO comparison table
loo_comp <- loo_compare(looQabkF_sub, loo_explicit)

loo_comp_table <- data.frame(
  Model     = c("K Model (estimated reward values)",
                "Explicit Model (fixed reward values)"),
  ELPD_diff = round(loo_comp[,1], 2),
  SE_diff   = round(loo_comp[,2], 2),
  ELPD_loo  = round(c(looQabkF_sub$estimates[1,1], loo_explicit$estimates[1,1]), 2),
  SE_loo    = round(c(looQabkF_sub$estimates[1,2], loo_explicit$estimates[1,2]), 2),
  p_loo     = round(c(looQabkF_sub$estimates[2,1], loo_explicit$estimates[2,1]), 2)
)

kable(loo_comp_table,
      caption = "LOO Model Comparison: K Model vs Explicit Ratings Model",
      col.names = c("Model", "ELPD Diff", "SE Diff", "ELPD LOO", "SE LOO", "p LOO")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(1, bold = TRUE, color = "white", background = "#2c7bb6")

# WAIC comparison table
waic_comp <- loo_compare(waic_k, waic_explicit)

waic_comp_table <- data.frame(
  Model     = c("K Model (estimated reward values)",
                "Explicit Model (fixed reward values)"),
  ELPD_diff = round(waic_comp[,1], 2),
  SE_diff   = round(waic_comp[,2], 2)
)

kable(waic_comp_table,
      caption = "WAIC Model Comparison: K Model vs Explicit Ratings Model",
      col.names = c("Model", "ELPD Diff", "SE Diff")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(1, bold = TRUE, color = "white", background = "#2c7bb6")
```

---


