---
title: "Portfolio 10"
author: "Anaelle Gackiere"
date: "`r Sys.Date()`"
---

## Part 1: Visualizing RCT Results of Streptomycin Therapy for Tuberculosis

I am interested in working in RCTs in the future, so I wanted to test out data visualization techniques that work well with two-arm trials. Alluvial diagrams are great for tracing patient flow through a trial and showing how individuals move from treatment assignment through intermediate outcomes to final results in a single graphic. For this piece I used the 1948 Streptomycin Trial, one of the first modern randomized controlled trials ever conducted, to explore both the overall treatment effect and who developed antibiotic resistance.

Source: `medicaldata` package, strep_tb data

```{r set-up, include=FALSE}
library("medicaldata")
library("ggalluvial") # for alluvial plot
library("ggplot2")
library("dplyr")
library("tidyverse")
library("fmsb") # for radar chart
library("ggridges")

```

```{r tb-prep}
data(strep_tb)
glimpse(strep_tb)

# data cleaning
strep_clean <- strep_tb %>%
  mutate(
    strep_resistance = recode(strep_resistance,
      "1_sens_0-8" = "Sensitive",
      "2_mod_8-99" = "Moderate",
      "3_resist_100+" = "Resistant"
    ),
    radiologic_6m = recode(radiologic_6m,
      "1_Death" = "Death",
      "2_Considerable_deterioration" = "Considerable\nDeterioration",
      "3_Moderate_deterioration" = "Moderate\nDeterioration",
      "4_No_change" = "No Change",
      "5_Moderate_improvement" = "Moderate\nImprovement",
      "6_Considerable_improvement" = "Considerable\nImprovement"
    ),
    baseline_condition = recode(baseline_condition,
      "1_Good" = "Good",
      "2_Fair" = "Fair",
      "3_Poor" = "Poor"
    ),
    baseline_temp = recode(baseline_temp,
      "1_98-98.9F" = "98-98.9°F",
      "2_99-99.9F" = "99-99.9°F",
      "3_100-100.9F" = "100-100.9°F",
      "4_100F+" = "100°F+"
    ),
    baseline_esr = recode(baseline_esr,
      "1_0-10" = "0-10",
      "2_11-20" = "11-20",
      "3_21-50" = "21-50",
      "4_51+" = "51+"
    ),
    baseline_cavitation = recode(baseline_cavitation,
      "0_no" = "No",
      "1_yes" = "Yes"
    )
  )

```

### Baseline Characteristics of TB sample.

The arms look well balanced across all characteristics given this was 1948 and randomization methods were primitive. Gender, baseline condition, cavitation, ESR all roughly equal between groups. One might notice a slight imbalance in that the streptomycin arm has slightly more patients in the highest fever category (100F+) than control.

```{r baseline-tb}

strep_table1 <- strep_clean %>%
  select(arm, gender, baseline_condition, baseline_temp, baseline_esr, baseline_cavitation) %>%
  rename(
    "Gender" = gender,
    "Baseline Condition" = baseline_condition,
    "Temperature" = baseline_temp,
    "ESR" = baseline_esr,
    "Cavitation on X-Ray" = baseline_cavitation
  ) %>%
  pivot_longer(cols = -arm, names_to = "variable", values_to = "value") %>%
  count(arm, variable, value) %>%
  group_by(arm, variable) %>%
  mutate(pct = n / sum(n))

ggplot(strep_table1, aes(x = value, y = pct, fill = arm)) +
  geom_col(position = "dodge", alpha = 0.85) +
  facet_wrap(~variable, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Streptomycin" = "plum4", "Control" = "pink")) +
  labs(
    title = "Baseline Characteristics by Treatment Arm",
    subtitle = "The 1948 Streptomycin Trial",
    x = NULL,
    y = "Percentage",
    fill = "Treatment Arm"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Alluvial Plot

```{r alluvial-plot}

strep_clean %>%
  group_by(arm, strep_resistance, radiologic_6m) %>%
  summarise(freq = n(), .groups = "drop") %>%
  ggplot(aes(axis1 = arm, 
             axis2 = strep_resistance, 
             axis3 = radiologic_6m,
             y = freq)) +
  geom_alluvium(aes(fill = arm), alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
  scale_fill_manual(values = c("Streptomycin" = "plum4", "Control" = "plum")) +
  labs(
    title = "Treatment Arm → Resistance → Outcome in the 1948 Streptomycin Trial",
    y = "Number of Patients",
    fill = "Treatment Arm"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

Streptomycin dominates the top outcomes (considerable improvement) while the control flows heavily toward death and deterioration. The historical finding was that all the resistance flows from the streptomycin arm only.

#### Who developed resistance?

Based on the plot below, resistance was not randomly distributed among streptomycin-treated patients. Those who developed resistance were disproportionately in poor baseline condition and more likely to present with cavitation on their chest X-ray which are indicators of more advanced disease. Higher ESR and fever followed the same pattern.

This interpretation was confirmed by follow-up analyses. Patients who developed resistance were more likely to present with cavitation (χ²(1) = 4.65, p = .031) and poor baseline condition (Fisher's exact, p \< .001) at enrollment. This suggests that higher disease severity may have given the bacterium more opportunity to mutate. This pattern remains relevant today, where the most vulnerable patients are often the most likely to develop antibiotic resistance

```{r step-resist}

strep_resist <- strep_clean %>%
  filter(arm == "Streptomycin") %>%
  filter(strep_resistance != "Moderate") # only 8 moderate cases, too few to compare
table(strep_resist$strep_resistance)

strep_resist %>%
  select(strep_resistance, baseline_condition, baseline_temp, 
         baseline_esr, baseline_cavitation) %>%
  pivot_longer(cols = -strep_resistance, 
               names_to = "variable", 
               values_to = "value") %>%
  count(strep_resistance, variable, value) %>%
  group_by(strep_resistance, variable) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = value, y = pct, fill = strep_resistance)) +
  geom_col(position = "dodge", alpha = 0.85) +
  facet_wrap(~variable, scales = "free_x", nrow = 2) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Sensitive" = "plum4", "Resistant" = "pink")) +
  labs(
    title = "Baseline Characteristics by Streptomycin Resistance",
    subtitle = "Who developed resistance in the Streptomycin arm?",
    x = NULL,
    y = "Percentage",
    fill = "Resistance Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r chi-square-tb}

# follow up chi square test for cavitation
strep_resist_binary <- strep_clean %>%
  filter(arm == "Streptomycin") %>%
  filter(strep_resistance %in% c("Sensitive", "Resistant")) %>%
  mutate(strep_resistance = droplevels(strep_resistance))

chisq.test(table(strep_resist_binary$baseline_cavitation, 
                 strep_resist_binary$strep_resistance))

# fisher for baseline condition
fisher.test(table(strep_resist_binary$baseline_condition, 
                  strep_resist_binary$strep_resistance))

```

## Part 2: Visualizing a Randomized Trial of Six Therapies for Scurvy

The `scurvy` dataset from the `medicaldata` package contains results from James Lind's 1757 trial testing six treatments for scurvy in 12 sailors. . I was drawn to it as a natural companion to the streptomycin piece since both are landmark historical trials that changed medicine, and both reward careful visualization despite their small sample sizes.

```{r scurvy-setup}
data(scurvy)
glimpse(scurvy)
```

```{r dataclean-scurvy}

radar_data <- scurvy %>%
  mutate(across(c(gum_rot_d6, skin_sores_d6, weakness_of_the_knees_d6, lassitude_d6),
                ~ as.numeric(str_extract(as.character(.), "\\d")))) %>%
  group_by(treatment) %>%
  summarise(across(c(gum_rot_d6, skin_sores_d6, weakness_of_the_knees_d6, lassitude_d6), mean)) %>%
  column_to_rownames("treatment")

# fmsb set up
radar_data <- rbind(
  max = rep(3, 4),
  min = rep(1, 4),
  radar_data
)

colnames(radar_data) <- c("Gum Rot", "Skin Sores", "Weak Knees", "Lassitude")

```

#### Scurvy Radar Chart

Since there are quite a few groups, it is difficult to read at first. Upon closer inspection though, one can see that citrus, in red, is at the center of the plot compared to all other treatments (and colors).

```{r}
colors <- c("yellow", "red", "green", "purple", "blue", "orange")

radarchart(radar_data,
  pcol = colors,
  pfcol = scales::alpha(colors, 0.2),
  plwd = 2,
  title = "Scurvy Symptoms by Treatment at Day 6"
)

legend("topright", legend = rownames(radar_data)[3:8],
       col = colors, lty = 1, lwd = 2, cex = 0.8)
```

#### Scurvy Heatmap

The heatmap better displays the stark difference in symptoms for citrus compared to the other treatments.

```{r}
scurvy_heat <- scurvy %>%
  mutate(across(c(gum_rot_d6, skin_sores_d6, weakness_of_the_knees_d6, lassitude_d6),
                ~ as.numeric(str_extract(as.character(.), "\\d")))) %>%
  group_by(treatment) %>%
  summarise(across(c(gum_rot_d6, skin_sores_d6, weakness_of_the_knees_d6, lassitude_d6), mean)) %>%
  pivot_longer(cols = -treatment, names_to = "symptom", values_to = "severity") %>%
  mutate(
    symptom = recode(symptom,
      "gum_rot_d6" = "Gum Rot",
      "skin_sores_d6" = "Skin Sores",
      "weakness_of_the_knees_d6" = "Weak Knees",
      "lassitude_d6" = "Lassitude"
    ),
    treatment = recode(treatment,
      "citrus" = "Citrus",
      "cider" = "Cider",
      "dilute_sulfuric_acid" = "Dilute Sulfuric Acid",
      "vinegar" = "Vinegar",
      "purgative_mixture" = "Purgative Mixture",
      "sea_water" = "Sea Water"
    )
  )

ggplot(scurvy_heat, aes(x = symptom, y = treatment, fill = severity)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(severity, 1)), color = "white", size = 4) +
  scale_fill_gradient(low = "plum", high = "plum4",
                      name = "Severity",
                      breaks = c(1, 2, 3),
                      labels = c("Mild", "Moderate", "Severe")) +
  labs(
    title = "Scurvy Symptom Severity by Treatment at Day 6",
    subtitle = "James Lind's 1757 trial [lower is better]",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())

```

James Lind's 1757 trial tested six treatments for scurvy aboard a Royal Navy vessel. The results are clearly unambiguous; citrus alone produced meaningful symptom relief across all four measures by day 6, while every other treatment left patients severely symptomatic.

## Part 3: Visualizing the Obstetrics and Periodontal Therapy RCT Dataset

Finally, I wanted to visualize a modern RCT using the OPT dataset, a trial examining whether treating maternal periodontal disease could reduce preterm birth risk. Given my interest in health equity, I was particularly curious whether treatment effects and birth outcomes differed by racial group.

Source: `medicaldata` package, opt data

Michalowicz et al., 'Treatment of periodontal disease and the risk of preterm birth', N Engl J Med 2006; 355:1885-1894. DOI: 10.1056/NEJMoa062249

```{r}
data(opt)
head(opt)
dim(opt)
```

#### Visualizing Birth Weight Outcomes

```{r opt-ridges}
opt_race <- opt %>%
  filter(!is.na(Birthweight)) %>%
  mutate(race = case_when(
    Black == "Yes" ~ "Black",
    White == "Yes" ~ "White",
    Hisp  == "Yes" ~ "Hispanic",
    Asian == "Yes" ~ "Asian",
    Nat.Am == "Yes" ~ "Native American",
    TRUE ~ "Other/Unknown"
  )) %>%
  filter(race != "Other/Unknown")

ggplot(opt_race, aes(x = Birthweight, y = race, fill = race)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 2500, linetype = "dashed", color = "gray40") +
  scale_fill_brewer(palette = "Set2") +
  annotate("text", x = 2500, y = 0.5, label = "Low birthweight\nthreshold (2500g)", 
           hjust = 1.1, size = 3.5, color = "gray40") +
  labs(
    title = "Birth Weight Distribution by Race",
    subtitle = "OPT Trial participants, periodontal therapy RCT",
    x = "Birth Weight (grams)",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

The Native American distribution looked strange, so I checked the sample size before making any further inferences:

```{r opt-table}

table(opt_race$race)
```

There are 7 Asian and 3 Native American participants. Those distributions are unreliable, so they have to be dropped.

```{r}
opt_race <- opt_race %>%
  filter(race %in% c("Black", "White", "Hispanic"))

ggplot(opt_race, aes(x = Birthweight, y = race, fill = race)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 2500, linetype = "dashed", color = "gray40") +
  scale_fill_brewer(palette = "Set2") +
  annotate("text", x = 2500, y = 0.5, label = "Low birthweight\nthreshold (2500g)", 
           hjust = 1.1, size = 3.5, color = "gray40") +
  labs(
    title = "Birth Weight Distribution by Race",
    subtitle = "OPT Trial participants, periodontal therapy RCT",
    x = "Birth Weight (grams)",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

Did periodontal treatment affect birthweight differently across racial groups?

```{r opt-ridges2}
ggplot(opt_race, aes(x = Birthweight, y = race, fill = Group)) +
  geom_density_ridges(alpha = 0.6, scale = 1.2) +
  geom_vline(xintercept = 2500, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("T" = "plum4", "C" = "plum"),
                    labels = c("T" = "Treatment", "C" = "Control")) +
  annotate("text", x = 2400, y = 0.5, label = "Low birthweight\nthreshold (2500g)", 
           hjust = 1.1, size = 3.5, color = "gray40") +
  labs(
    title = "Birth Weight Distribution by Race and Treatment Arm",
    subtitle = "Did periodontal treatment affect birthweight differently across racial groups?",
    x = "Birth Weight (grams)",
    y = NULL,
    fill = "Treatment Arm"
  ) +
  theme_minimal(base_size = 13)
```

Periodontal treatment did not meaningfully shift birthweight distributions within any racial group. The treatment and control curves overlap substantially for Black, Hispanic, and White participants alike. This confirms the trial's published null finding that treating maternal periodontal disease did not reduce preterm birth or low birthweight risk regardless of race.

However the racial disparity in birthweight is clearly visible regardless of treatment arm. Black participants show a left-shifted distribution with more mass below the 2500g low birthweight threshold compared to White and Hispanic participants. Periodontal treatment did nothing to close that gap.

#### Visualizing Preterm Births

```{r opt-preterm}
opt_race <- opt_race %>%
  mutate(Preg.ended...37.wk = trimws(as.character(Preg.ended...37.wk)))

opt_race %>%
  filter(Preg.ended...37.wk %in% c("Yes", "No")) %>%
  group_by(race, Group, Preg.ended...37.wk) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(race, Group) %>%
  mutate(pct = n / sum(n)) %>%
  filter(Preg.ended...37.wk == "Yes") %>%
  ggplot(aes(x = race, y = pct, fill = Group)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c("T" = "plum4", "C" = "plum"),
                    labels = c("T" = "Treatment", "C" = "Control")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Preterm Birth Rate by Race and Treatment Arm",
    x = NULL,
    y = "Preterm Birth Rate",
    fill = "Treatment Arm"
  ) +
  theme_minimal(base_size = 13)
```

Black participants had the highest preterm birth rates overall, and treatment made virtually no difference. Hispanic participants showed the largest gap between arms, with the treatment group having lower preterm birth rates, though small sample sizes make this hard to interpret confidently. White participants showed the opposite pattern, with slightly higher rates in the treatment arm.

The overall null finding might mask potential heterogeneity by race. The intervention may have had differential effects that were washed out in the aggregate analysis.

##### Logistic Regression with interaction term between race and treatment group

```{r logreg-preterm}
preterm_model <- glm(
  (Preg.ended...37.wk == "Yes") ~ race * Group,
  data = opt_race,
  family = binomial
)

summary(preterm_model)
```

The logistic regression with race by treatment interaction terms found no statistically significant differential treatment effects by race (all p \> .40), which suggests the visual differences observed are likely due to chance given limited statistical power. Adequately powered subgroup analyses would require a much larger trial with prespecified racial subgroup analyses which is a design consideration that remains underutilized in RCTs today.
